Eco-evolutionary dynamics of finite populations from first principles

### 2023-03-10 ind_files/02_BD_processes.tex
Explain that FPE is a continuity equation

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004} and pure mathematics~\citep{oksendal_stochastic_1998} literature. If the function $G$ is independent of $x$, then it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

>>>>

I present a simple informal derivation in Appendix \ref{App_SDE_FPE} for the sake of completeness. Equation \eqref{ito_FPE} is called the `Fokker-Planck equation' in the physics and applied mathematics literature~\citep{gardiner_stochastic_2009} and is often called the `Kolmogorov forward equation' in the population genetics~\citep{ewens_mathematical_2004} and pure mathematics~\citep{oksendal_stochastic_1998} literature. As I explain in Appendix \ref{App_SDE_FPE}, the Fokker-Planck equation is a `conservation law' for probability, and just mathematically expresses the common-sense observation that the sum (or integral) of probabilities over the entire state space cannot change over time (since it must always equal 1). If the function $G$ is independent of $x$, it comes out of the derivatives in equation \eqref{ito_FPE}, and the resultant Fokker-Planck equation is said to be `linear' (and is much easier to solve). This link between SDEs and Fokker-Planck equations goes both ways: One can show that every stochastic process with a probability density described by a Fokker-Planck equation of the form \eqref{ito_FPE} corresponds to the solution of an SDE of the form \eqref{ito_SDE_diff}, though the proof is much more technical and will not be discussed here. This two-way correspondence proves to be extremely useful, as one approach often works for applications in which the other fails. This correspondence makes it greatly desirable to be able to describe our stochastic process of interest as either the solution to an It\^o SDE of the form \eqref{ito_SDE_diff} or as the solution to a Fokker-Planck equation of the form \eqref{ito_FPE}. System-size expansions facilitate such a description for birth-death processes.

### 2023-03-10 ind_files/99_Appendix_A.tex
Explain that FPE is a continuity equation

Equation \eqref{FPE} is the Fokker-Planck equation in one dimension. Using the exact same strategy, the multidimensional Fokker-Planck equation for the $n$ dimensional It\^{o} Process $d\mathbf{X}_t = \mu(\mathbf{X}_t,t)dt + \sigma(\mathbf{X}_t,t)dB_t$ is found to be:

@@

\label{FPE_ndim}
\frac{\partial P}{\partial t}(\mathbf{x},t) =-\sum\limits_{i=1}^{n}\frac{\partial}{\partial x_i}\left(\mu_i(\mathbf{x},t) P(\mathbf{x},t)\right) + \frac{1}{2}\sum\limits_{i=1}^{n}\sum\limits_{j=1}^{n}\frac{\partial^2}{\partial x_ix_j}\left(D_{ij}P(\mathbf{x},t)\right)

@@

where $\mathbf{D} = \mathbf{\sigma}\mathbf{\sigma}^T$.

>>>>

Equation \eqref{FPE} is the Fokker-Planck equation in one dimension. Using the exact same strategy, the multidimensional Fokker-Planck equation for the $m$-dimensional It\^{o} Process $d\mathbf{X}_t = \boldsymbol{\mu}(\mathbf{X}_t,t)dt + \boldsymbol{\sigma}(\mathbf{X}_t,t)dB_t$ can be easily found to be:

@@

\label{app_FPE_ndim}
\frac{\partial P}{\partial t}(\mathbf{x},t) =-\sum\limits_{i=1}^{m}\frac{\partial}{\partial x_i}\left(\mu_i(\mathbf{x},t) P(\mathbf{x},t)\right) + \frac{1}{2}\sum\limits_{i=1}^{m}\sum\limits_{j=1}^{m}\frac{\partial^2}{\partial x_i \partial x_j}\left(D_{ij}P(\mathbf{x},t)\right)

@@

where $\mathbf{D} = \mathbf{\sigma}\mathbf{\sigma}^T$. As an aside, I briefly note that some authors like to define the `probability current' $\mathbf{J}(\mathbf{x},t)$, an $m$-dimensional function with $i\textsuperscript{th}$ element
J_i(\mathbf{x},t) \coloneqq \mu_i(\mathbf{x},t)P(\mathbf{x},t) - \frac{1}{2}\sum\limits_{j=1}^{m}\frac{\partial}{\partial x_j}\left(D_{ij}P(\mathbf{x},t)\right)
In this notation, equation \eqref{app_FPE_ndim} can be written in the more compact form:
\label{app_FPE_ndim_continuity_eqn}
\frac{\partial P}{\partial t}  + \div{\mathbf{J}} = 0
where $\div$ is the divergence operator (also denoted $\mathrm{div}$ in some texts). Those familiar with physics should immediately recognize that equation \eqref{app_FPE_ndim_continuity_eqn} is in the form of a so-called `continuity equation' for a conserved quantity. Continuity equations occur in various areas of applied mathematics, most famously in electromagnetism and fluid dynamics. This explains the name `probability current', in analogy with other currents such as electrical current or fluid current. The continuity equation representation also  makes it clear that the Fokker-Planck equation describes the `flow of probability' in the system. In particular, equation \eqref{app_FPE_ndim_continuity_eqn} says that the total probability in the system is `conserved', and is thus simply a mathematical formalization of the common-sense idea that whenever the probability of the system of being in a given state decreases, the probability of it being in some other state must increase (and vice versa).
